### 2024-08-15 machine-learning.qmd
20240815 - tree-boosting QBs

@@

```{r}
# Gamma distribution requires positive-only values
data_train_qb_matrix[,"fantasyPoints_lag"][data_train_qb_matrix[,"fantasyPoints_lag"] <= 0] <- 0.01

# Predictor variables
pred_vars_qb <- data_train_qb_matrix %>% 
  as_tibble() %>% 
  select(-gsis_id, -fantasyPoints_lag, -ageCentered20, ageCentered20Quadratic) %>% 
  names()

#pred_vars_qb_categorical <- "" # to specify categorical predictors

gp_model_qb <- GPModel(
  group_data = data_train_qb_matrix[,"gsis_id"],
  likelihood = "gamma", # "gaussian"
  group_rand_coef_data = cbind(
    data_train_qb_matrix[,"ageCentered20"],
    data_train_qb_matrix[,"ageCentered20Quadratic"]),
  ind_effect_group_rand_coef = c(1,1))

gp_data_qb <- gpb.Dataset(
  data = data_train_qb_matrix[,pred_vars_qb],
  label = data_train_qb_matrix[,"fantasyPoints_lag"])

#gpb.Dataset.set.categorical(gp_data_qb, pred_vars_qb_categorical) # to specify categorical predictors

params_qb <- list(
  learning_rate = 0.01,
  max_depth = 2,
  num_leaves = 2^10,
  min_data_in_leaf = 10,
  lambda_l2 = 10,
  num_threads = num_cores)

nrounds_qb <- 379

gp_model_qb$set_optim_params(params = list(optimizer_cov = "nelder_mead"))

gp_model_fit_qb <- gpb.train(
  data = gp_data_qb,
  gp_model = gp_model_qb,
  nrounds = nrounds_qb,
  params = params_qb) # verbose = 0

summary(gp_model_qb) # Estimated random effects model

# Test Model on Test Data
pred_test_qb <- predict(
  gp_model_fit_qb,
  data = data_test_qb_matrix[,pred_vars_qb],
  group_data_pred = data_test_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    data_test_qb_matrix[,"ageCentered20"],
    data_test_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

y_pred_test_qb <- pred_test_qb[["response_mean"]]
mean((data_test_qb_matrix[,"fantasyPoints_lag"] - y_pred_test_qb)^2)

# Generate model predictions for next season
pred_nextYear_qb <- predict(
  gp_model_fit_qb,
  data = newData_qb_matrix[,pred_vars_qb],
  group_data_pred = newData_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    newData_qb_matrix[,"ageCentered20"],
    newData_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

@@

newData_qb %>% 
  arrange(-fantasyPoints_lag) %>% 
  select(name, fantasyPoints_lag)
```

>>>>

### Specify Options {#machineLearningSpecifyOptions}

```{r}
options(scipen = 999) # prevent scientific notation
```

@@

#### Process Data {#sec-treeBoostingProcessData}

If using a gamma distribution, it requires positive-only values:

```{r}
#data_train_qb_matrix[,"fantasyPoints_lag"][data_train_qb_matrix[,"fantasyPoints_lag"] <= 0] <- 0.01
```

#### Specify Predictor Variables {#sec-treeBoostingSpecifyPredictors}

```{r}
pred_vars_qb <- data_train_qb_matrix %>% 
  as_tibble() %>% 
  select(-fantasyPoints_lag, -ageCentered20, ageCentered20Quadratic) %>% # -gsis_id
  names()

pred_vars_qb_categorical <- "gsis_id" # to specify categorical predictors
```

#### Specify General Model Options {#sec-treeBoostingSpecifyGeneralModelOptions}

```{r}
model_likelihood <- "gaussian" # gamma
nrounds <- 1000
```

#### Identify Optimal Tuning Parameters {#sec-treeBoostingOptimalTuningParameters}

```{r}
# Partition training data into inner training data and validation data
ntrain_qb <- dim(data_train_qb_matrix)[1]

set.seed(52242)
valid_tune_idx_qb <- sample.int(ntrain_qb, as.integer(0.2*ntrain_qb))

folds_qb <- list(valid_tune_idx_qb)

# Specify parameter grid, gp_model, and gpb.Dataset
param_grid_qb <- list(
  "learning_rate" = c(1,0.1,0.01),
  "max_depth" = c(1,2,3,5,10),
  "min_data_in_leaf" = c(10,100,1000),
  "lambda_l2" = c(0,1,10))

other_params_qb <- list(num_leaves = 2^10)

gp_model_qb <- GPModel(
  group_data = data_train_qb_matrix[,"gsis_id"],
  likelihood = model_likelihood,
  group_rand_coef_data = cbind(
    data_train_qb_matrix[,"ageCentered20"],
    data_train_qb_matrix[,"ageCentered20Quadratic"]),
  ind_effect_group_rand_coef = c(1,1))

gp_data_qb <- gpb.Dataset(
  data = data_train_qb_matrix[,pred_vars_qb],
  categorical_feature = pred_vars_qb_categorical,
  label = data_train_qb_matrix[,"fantasyPoints_lag"])

# Find optimal tuning parameters
opt_params_qb <- gpb.grid.search.tune.parameters(
  param_grid = param_grid_qb,
  params = other_params_qb,
  num_try_random = NULL,
  folds = folds_qb,
  data = gp_data_qb,
  gp_model = gp_model_qb,
  nrounds = nrounds,
  early_stopping_rounds = 10,
  verbose_eval = 1,
  metric = "mse")

opt_params_qb
```

#### Specify Model and Tuning Parameters {#sec-treeBoostingSpecifyModel}

```{r}
gp_model_qb <- GPModel(
  group_data = data_train_qb_matrix[,"gsis_id"],
  likelihood = model_likelihood,
  group_rand_coef_data = cbind(
    data_train_qb_matrix[,"ageCentered20"],
    data_train_qb_matrix[,"ageCentered20Quadratic"]),
  ind_effect_group_rand_coef = c(1,1))

gp_data_qb <- gpb.Dataset(
  data = data_train_qb_matrix[,pred_vars_qb],
  categorical_feature = pred_vars_qb_categorical,
  label = data_train_qb_matrix[,"fantasyPoints_lag"])

params_qb <- list(
  learning_rate = 1,
  max_depth = 10,
  num_leaves = 2^10,
  min_data_in_leaf = 10,
  lambda_l2 = 0,
  num_threads = num_cores)

#gp_model_qb$set_optim_params(params = list(optimizer_cov = "nelder_mead")) # to speed up model estimation
```

#### Fit Model {#sec-treeBoostingFitModel}

```{r}
gp_model_fit_qb <- gpb.train(
  data = gp_data_qb,
  gp_model = gp_model_qb,
  nrounds = nrounds,
  params = params_qb) # verbose = 0

summary(gp_model_qb) # Estimated random effects model
```

#### Evaluate Accuracy of Model on Test Data {#sec-treeBoostingModelAccuracy}

```{r}
# Test Model on Test Data
pred_test_qb <- predict(
  gp_model_fit_qb,
  data = data_test_qb_matrix[,pred_vars_qb],
  group_data_pred = data_test_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    data_test_qb_matrix[,"ageCentered20"],
    data_test_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

y_pred_test_qb <- pred_test_qb[["response_mean"]]
cbind(y_pred_test_qb, data_test_qb_matrix[,"fantasyPoints_lag"])
mean((data_test_qb_matrix[,"fantasyPoints_lag"] - y_pred_test_qb)^2)
```

#### Generate Predictions for Next Season {#sec-treeBoostingModelPredictions}

```{r}
# Generate model predictions for next season
pred_nextYear_qb <- predict(
  gp_model_fit_qb,
  data = newData_qb_matrix[,pred_vars_qb],
  group_data_pred = newData_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    newData_qb_matrix[,"ageCentered20"],
    newData_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

@@

newData_qb %>% 
  arrange(-fantasyPoints_lag) %>% 
  select(name, fantasyPoints_lag, fantasy_points)
```
